---
title: "Getting Started with spatialPointsEditor"
author: "Dr. Alavi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spatialPointsEditor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(spatialPointsEditor)
library(sf)
```

## Introduction

The **spatialPointsEditor** package provides interactive tools to visualize and correct coordinates of `sf` POINT objects in R. This vignette demonstrates how to use the main functions to edit spatial point data with visual feedback.

## Why Use This Package?

Sometimes spatial point data has incorrect coordinates due to:

- GPS errors during field data collection
- Geocoding mistakes
- Manual data entry errors
- Points falling outside expected boundaries
- Coordinate system transformation issues

The **spatialPointsEditor** makes it easy to identify and correct these errors interactively.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("GeoLearnR/spatialPointsEditor")
```

## Quick Start Example

### Load Sample Data

The package includes example data:

```{r}
# Load example points
data(example_points)

# View the data
print(example_points)
```

### View Points on Map

Before editing, let's visualize the points:

```{r eval=FALSE}
library(mapview)
mapview(example_points)
```

### Edit Points Interactively

Launch the interactive editor:

```{r eval=FALSE}
# Edit the points
corrected_points <- edit_spatial_points(
  sf_data = example_points,
  id_column = "id"
)
```

The editor window will open with:

- **Map View**: Interactive map showing your points
- **Data Table**: Editable table of coordinates
- **Modification Log**: Track all changes
- **Help Tab**: Detailed instructions

### How to Edit

There are **three ways** to edit coordinates:

#### Method 1: Drag and Drop on Map

1. Click a point to select it (turns red)
2. Drag the red marker to the correct location
3. Changes are saved automatically

#### Method 2: Edit in Data Table

1. Go to the "Data Table" tab
2. Double-click a longitude or latitude cell
3. Enter the correct value
4. Press Enter

#### Method 3: Manual Entry

1. Click a point to select it
2. In the sidebar, enter new coordinates
3. Click "Apply Coordinates"

### Compare Changes

After editing, compare the original and corrected data:

```{r eval=FALSE}
# Compare original vs edited points
comparison <- compare_edits(
  original = example_points,
  edited = corrected_points,
  id_column = "id"
)

# View the comparison
print(comparison)

# Print summary statistics
print_edit_summary(comparison)
```

The comparison shows:

- Original and new coordinates
- Distance moved for each point
- Which points were modified

## Real-World Example

### Loading Your Own Data

```{r eval=FALSE}
# Load shapefile
my_points <- st_read("data/field_sites.shp")

# Or load from other formats
my_points <- st_read("data/locations.gpkg")
my_points <- st_read("data/points.geojson")
```

### Complete Workflow

```{r eval=FALSE}
# 1. View original data
mapview(my_points)

# 2. Identify incorrect points visually

# 3. Edit interactively
corrected <- edit_spatial_points(my_points, id_column = "site_id")

# 4. Review changes
comparison <- compare_edits(my_points, corrected, id_column = "site_id")
print_edit_summary(comparison)

# View only modified points
modified_only <- comparison[comparison$Modified, ]
print(modified_only)

# 5. Save corrected data
st_write(corrected, "data/field_sites_corrected.shp", delete_dsn = TRUE)

# 6. Export change log
write.csv(comparison, "data/corrections_log.csv", row.names = FALSE)
```

## Editor Features

### Undo/Redo

- Click **"↶ Undo"** to revert changes
- Click **"↷ Redo"** to reapply changes
- Unlimited undo/redo history

### Reset Points

- **Reset Selected Point**: Revert one point to original coordinates
- **Reset All Points**: Revert all points to original coordinates

### Modification Tracking

The **Modification Log** tab shows:

- Original coordinates
- New coordinates
- Change in longitude/latitude
- Distance moved (in meters)

Download the log as CSV for documentation.

### Visual Feedback

Points are color-coded:

- 🔵 **Blue**: Original, unmodified points
- 🟠 **Orange**: Modified points
- 🔴 **Red**: Currently selected point (draggable)

## Advanced Usage

### Return Only Modified Points

```{r eval=FALSE}
# Get only the points that were changed
modified_only <- edit_spatial_points(
  sf_data = my_points,
  id_column = "id",
  return_modified = TRUE
)
```

### Working with Different CRS

The editor automatically handles coordinate reference systems:

```{r eval=FALSE}
# Data in UTM Zone 33N
utm_points <- st_transform(example_points, crs = 32633)

# Edit (automatically converts to WGS84 for display)
edited_utm <- edit_spatial_points(utm_points, id_column = "id")

# Result is returned in original CRS (UTM 33N)
st_crs(edited_utm)
```

### Batch Editing Multiple Datasets

```{r eval=FALSE}
# Edit multiple related datasets
sites <- edit_spatial_points(sites_sf, id_column = "site_id")
plots <- edit_spatial_points(plots_sf, id_column = "plot_id")
trees <- edit_spatial_points(trees_sf, id_column = "tree_id")
```

## Tips and Best Practices

### Before Editing

1. **Backup your data** before making changes
2. **Document why** corrections are needed
3. **Use meaningful IDs** to track points

### During Editing

1. **Zoom in** on the map for precision
2. **Use satellite imagery** to verify locations
3. **Check the modification log** regularly
4. **Use undo** if you make a mistake

### After Editing

1. **Review all changes** in the comparison table
2. **Save the change log** for documentation
3. **Verify** corrected points make sense
4. **Document** the editing process in your workflow

## Common Use Cases

### 1. Correcting GPS Errors

Field-collected GPS points often have accuracy issues:

```{r eval=FALSE}
# Field data with GPS errors
field_data <- st_read("gps_points.gpkg")

# Edit to align with actual features (roads, buildings)
corrected_field <- edit_spatial_points(field_data, id_column = "point_id")
```

### 2. Fixing Geocoding Mistakes

Geocoded addresses may be incorrect:

```{r eval=FALSE}
# Geocoded locations
addresses <- st_read("geocoded_addresses.shp")

# Manually correct misplaced points
fixed_addresses <- edit_spatial_points(addresses, id_column = "address_id")
```

### 3. Adjusting Points to Boundaries

Ensure points fall within expected boundaries:

```{r eval=FALSE}
# Plot centers that should be inside forest boundary
plot_centers <- st_read("plot_centers.shp")

# Adjust points that fall outside
adjusted_plots <- edit_spatial_points(plot_centers, id_column = "plot_num")
```

## Troubleshooting

### Editor Won't Open

```{r eval=FALSE}
# Make sure all dependencies are installed
install.packages(c("shiny", "leaflet", "sf", "DT", "shinyjs"))
```

### Points Not Showing on Map

Check your CRS:

```{r eval=FALSE}
# Check CRS
st_crs(my_points)

# If missing, set it
st_crs(my_points) <- 4326  # WGS84
```

### Can't Edit Specific Columns

Only longitude and latitude can be edited in the data table. Other attributes are read-only.

## Getting Help

- **Package documentation**: `?edit_spatial_points`
- **GitHub issues**: https://github.com/GeoLearnR/spatialPointsEditor/issues
- **Help tab** in the editor interface

## Citation

If you use this package in your research, please cite:

```{r eval=FALSE}
citation("spatialPointsEditor")
```

## Summary

The **spatialPointsEditor** package makes it easy to:

✅ Visually identify incorrect coordinates  
✅ Edit points interactively with drag-and-drop  
✅ Track all modifications with undo/redo  
✅ Export detailed change logs  
✅ Work with any CRS  

For more examples and updates, visit: https://github.com/GeoLearnR/spatialPointsEditor
```
